### 2025-04-08 correlation.qmd
20250408 - additions

The $r$ value (`{r} petersenlab::apa(pearson_r, decimals = 2)`) is slightly negative.
The [effect size](#sec-practicalSignificance) of the association is small—for what is considered a small, medium, and large effect size for $r$ values, see @sec-effectSizeThresholds.
The $p$-value is less than .05, so it is a statistically significant correlation.
We could report this finding as follows: There was a statistically significant, weak negative association between height and fantasy points ($r[`{r} pearson_df`] = `{r} petersenlab::apa(pearson_r, decimals = 2)`$, $p < .001$); the taller a player was, the fewer fantasy points they tended to score.

@@

Please note that the $p$-value for a correlation assumes that the observations are independent—in particular, that the residuals are not correlated.
However, the observations are not independent in the `player_stats_seasonal` dataframe used above, because the same player has multiple rows—one row corresponding to each season they played.
This non-independence violates the traditional assumptions of the significance test of a correlation.
We could address this assumption by analyzing only one season from each player.
However, for simplicity, we present results from the whole dataframe.
In @sec-mixedModels, we discuss [mixed model](#sec-mixedModels) approaches that handle repeated measures and other data that violate assumptions of non-independence that are shared by correlation and [multiple regression](#sec-multipleRegression); assumptions of [multiple regression](#sec-multipleRegression) are described in @sec-assumptionsRegression.

### Spearman Correlation {#sec-correlationExamplesSpearman}

@@

@@

@@

@@

>>>>

The $r$ value (`{r} petersenlab::apa(pearson_r, decimals = 2, leading = FALSE)`) is slightly negative.
The [effect size](#sec-practicalSignificance) of the association is small—for what is considered a small, medium, and large effect size for $r$ values, see @sec-effectSizeThresholds.
The $p$-value is less than .05, so it is a statistically significant correlation.
We could report this finding as follows: There was a statistically significant, weak negative association between height and fantasy points ($r(`r pearson_df`) = `r petersenlab::apa(pearson_r, decimals = 2, leading = FALSE)`$, $p `r petersenlab::pValue(pearson_p)`$); the taller a player was, the fewer fantasy points they tended to score.

@@

### Rank Correlation: Spearman's Rho ($\rho$) {#sec-correlationExamplesSpearman}

Spearman's rho ($\rho$) is a rank correlation.
Spearman's rho does not assume that the data are [interval](#sec-interval) or [ratio](#sec-ratio) level of measurement.
Unlike [Pearson correlation](sec-correlationExamplesPearson), Spearman's rho allows estimating associations among variables that are [ordinal](#sec-ordinal) level of measurement.

For instance, using Spearman's rho, we could examine the association between a player's draft round and their fantasy points, because draft round is an [ordinal](#sec-ordinal) variable:

```{r}
cor(
  player_stats_seasonal[,c("draftround","fantasyPoints")],
  use = "pairwise.complete.obs",
  method = "spearman")

cor.test(
  ~ draftround + fantasyPoints,
  data = player_stats_seasonal,
  method = "spearman"
)
```

spearman_test <- cor.test(
  ~ draftround + fantasyPoints,
  data = player_stats_seasonal,
  method = "spearman"
)

spearman_r <- spearman_test$estimate
spearman_df <- cor.test(
  ~ draftround + fantasyPoints,
  data = player_stats_seasonal)$parameter
spearman_p <- spearman_test$p.value
```

Using Spearman's rho, there is a negative association between draft round and fantasy points ($r(`r spearman_df`) = `r petersenlab::apa(spearman_r, decimals = 2, leading = FALSE)`$, $p `r petersenlab::pValue(spearman_p)`$); unsurprisingly, the earlier the round a player is drafted, the more fantasy points they tend to score.

### Partial Correlation {#sec-correlationExamplesNonlinear}

@@

```{r}
set.seed(52242) # for reproducibility

XICOR::xicor(
  player_stats_seasonal$height,
  player_stats_seasonal$fantasyPoints,
  pvalue = TRUE
)

set.seed(52242) # for reproducibility

XICOR::calculateXI(
  player_stats_seasonal$height,
  player_stats_seasonal$fantasyPoints)
```

@@

The [`petersenlab`](https://cran.r-project.org/web/packages/petersenlab/index.html) package [@R-petersenlab] contains the `cor.table()` function that generates a correlation matrix of variables, including the $r$ value, number of observations ($n$), and $p$-value for each association.
Here is a Pearson correlation matrix:

```{r}
petersenlab::cor.table(
  player_stats_seasonal[,c("height","weight","fantasyPoints")]
)
```

Here is a Spearman correlation matrix:

```{r}
petersenlab::cor.table(
  player_stats_seasonal[,c("height","weight","fantasyPoints")],
  correlation = "spearman"
)
```

Here is a correlation matrix of variables that includes just the $r$ value and asterisks denoting the statistical significance of the association, for greater concision: $^\dagger p < .1$; $^* p < .05$; $^{**} p < .01$; $^{***} p < .001$.

```{r}
petersenlab::cor.table(
  player_stats_seasonal[,c("height","weight","fantasyPoints")],
  type = "manuscript" # include r values and asterisks
)
```

Here is a correlation matrix of variables that includes just the $r$ value for even greater concision.

```{r}
petersenlab::cor.table(
  player_stats_seasonal[,c("height","weight","fantasyPoints")],
  type = "manuscriptBig" # include just r values
)
```

Here is a partial correlation matrix, controlling for the player's age:

```{r}
petersenlab::partialcor.table(
  player_stats_seasonal[,c("height","weight","fantasyPoints")],
  z = player_stats_seasonal$age # control for the player's age
)
```

@@

## Addressing Non-Independence of Observations {#sec-correlationNonIndependence}

Please note that the $p$-value for a correlation assumes that the observations are independent—in particular, that the residuals are not correlated.
However, the observations are not independent in the `player_stats_seasonal` dataframe used above, because the same player has multiple rows—one row corresponding to each season they played.
This non-independence violates the traditional assumptions of the significance test of a correlation.
We could address this assumption by analyzing only one season from each player or by estimating the significance of the correlation coefficient using cluster-robust standard errors.
For simplicity, we present results above from the whole dataframe.
In @sec-mixedModels, we discuss [mixed model](#sec-mixedModels) approaches that handle repeated measures and other data that violate assumptions of non-independence that are shared by correlation and [multiple regression](#sec-multipleRegression); assumptions of [multiple regression](#sec-multipleRegression) are described in @sec-assumptionsRegression.
Below, we demonstrate how to account for non-independence of observations using cluster-robust standard errors.

```{r}

```

@@

### Examining Robust Correlation {#sec-correlationRobust}
